Enzo: An Adaptive Mesh Refinement Code for Astrophysics

### 2013-05-29 parallel.tex
further edits to parallel section.

The current version of \enzo\ has been parallelized for distributed
memory platforms using the Message Passing Interface (MPI).  This is
done using a single grid object as the basic unit of parallelization.
Each grid object -- including all cell and grid data -- is fully
contained in a single processor\footnote{We use the word {\it
processor} to mean a basic distribution unit; this could by either a
core or a node, depending on details of the system.  Note that the
current version of Enzo is not threaded, although the development of a
hybrid version is in process}.  Parallelization is accomplished by
distributing grids amongst processors.  This is done on the root grid
using a simple tiling system, where the root grid is split up into
$N_{\rm root}$ tiles, where $N_{\rm root}$ is typically equal to the
number of processors $N_p$.

Load balancing on levels other than the root level (i.e., grids for
which the level $l > 0$) is different as the refined patches are not
generally uniformly distributed.  Grids on refined levels are first
placed on the same processor as their parent to minimize
communication; however, this generally does not result in a
well-balanced load.  Therefore, the code has a number of options for
load balancing the grids on a given level. Each grid is assigned a
load, generally equal to the number of cells in the grid (which,
empirically, is a good estimate of computational cost).  The first
load-balancing option is to move a grid from the processor with the
highest load to the processor with the lowest load, with the proviso
that only grids with load factors less than half the difference
between the highest and lowest loaded processors will be moved.  This
continues until the load ratio between the most-to-least loaded
processor is below 1.05 or until no suitable grid can be found to
transfer.  A second load balancing option uses a space-filling Hilbert
curve to order the grids by their approximate spatial position.  Then,
once the grids have a specific one-dimensional ordering, we can divide
up the grids into $N_p$ groups (with the division taking place as
equally as possible).  Load balancing is done separately for each
level.  Clearly load balancing is most successful if there are
significantly more grids than processors; however, small grids are
less efficient (because of their ghost zones), and so the code uses a
simple heuristic in order to split up grids until there are of order
10 grids per processor.  This generally results in good load-balancing
while not producing grids that are wastefully small.

@@

Because of the wide variety of simulations, methods, and uses of Enzo,
it is difficult to state generally which routines will be most costly
during a given simulation.  As such, we have designed a lightweight
registering system that has been implemented for the most commonly
used routines (such as the hydrodynamic and gravity solvers) as well
as refinement level timers that measure the time spent on each level.
Beyond this minimal set of routines, we have designed a simple way for
the user to modify the source by adding \texttt{TIMER\_START(``Your
Routine Name'')} and \texttt{TIMER\_END(``Your Routine Name'')}.
These timers are created and managed individually on each processor in
an asynchronous fashion, and contribute minimal computational, memory,
and IO overhead.

At each complete root grid time step (or less often if specified),
each timer is then communicated to the root processor where it
calculates the mean, standard deviation, minimum, and maximum for each
of the timers across all processors.  For level timers, there are also
attributes such as the number of cell updates, the current number of
grids, and the average cells/s/MPI process.  This information is then
output to a logfile.  This provides a simplified interface to the user
that can be used to diagnose performance issues as well as estimate a
given problem type's scalability.  In addition to the logfile, we have
developed a plotting interface for quickly producing figures that
process the data from the logfile.  These capabilities are described
in the online documentation, along with a further discussion of the
performance measurement implementation.

@@

In this section, we develop a simple, approximate model to estimate
the compute time and memory required to complete a given calculation.
Note that we are {\it not} trying to model parallel performance (which
was discussed in the previous section), but have an even simpler goal:
to determine how to scale compute time estimates as we increase the simulation
resolution.  This is a straightforward thing to do in codes that use
static grids, as the computational effort per time step is constant.
However, for an AMR calculation, the number of grids that will be
generated during the run is not known in advance, and so the CPU time
per problem time can vary drastically throughout the calculation.
For example, at the beginning of the simulation when the densities are nearly uniform,
only the static grid is required and the calculation progresses
rapidly.  However, as structure forms and dense clumps are generated,
the number of grid points swells by orders of magnitude (an increase
of $10^3$ is not uncommon) and most of the cpu time is consumed at
late times.  Therefore, simply performing a few steps does produce a
good estimate of the cpu time.

@@

To make progress, we assume that the computational cost to advance a
single cell by one timestep is a constant.  This can be incorrect if
the chemistry solver requires many iterations, but is usually fairly
accurate.  For a unigrid calculation, the time would by proportional
to $N_{\rm root}^4$ since the number of cells scales as $N_{\rm
root}^3$ and, assuming the Courant condition is the time limiting
factor, then the number of steps to advance a given time is
proportional to $N_{\rm root}.$ Therefore, to advance a hierarchy a
given time, we find, accounting for all levels and using a refinement
factor of 2.
\begin{equation} t_{\rm SU} = C_1 \sum_l f_l N_{\rm root}^4 2^{4l}
\end{equation} where $C_1$ is a constant, which can be thought of as
the time taken to advance a single cell.  The factor $f_l$ is the
fraction of the volume on a given level that is actually refined.  By
definition $f_0 = 1$, and $f_l \le 1$.  In writing down this equation,
we neglect a number of costs which are not directly proportional to
cell count, including communication between processors, the $\log{N}$
factor for the root grid FFT, Cache misses, optimizations, and other
costs associated with processing the hierarchy.  The first item on
this list, in particular, is clearly not true for large processor
counts; however, we neglect parallel considerations in this section.

@@

\begin{figure}
\centerline{\includegraphics[width=0.7\textwidth]{figures/scaling_plot2}}
\caption{This figure show (clockwise from upper left), the covering
fraction $f_l$ of the grids on a given level $l$, the ratio of
$f_{l}/f_{l+1}$, the relative memory usage of each level (normalized
by the memory usage of level 0), and, in the lower-left panel, the
relative cpu usage of each level, again normalized by $l=0$.  In each
panel, three curves are shown: the solid black and long dashed red are
two cosmological simulations with box sizes of 30 h$^{-1}$ Mpc and
80h$^{-1}$ Mpc respectively.  The dot-dashed blue line is for a
non-cosmological simulation of a ram-pressure stripped galaxy.  In the
top-right panel, there is a line at the critical ratio of 0.06, which
determines if the next finer level ($l+1)$ takes more (above) or less
(below) cpu time than level $l$.}
\label{fig:scaling}
\end{figure}

As, the figure demonstrates, the $f_l$ values all show a sharp decline
with level, dropping as power-laws with the level $l$.  Focusing first
on the ``refine-everywhere" cosmological runs, which both show similar
behavior despite the different box sizes and redshifts at which we
collect the data.  In fact, we find the results are very typical for
cosmological runs, with only slight variations depending on if we are
simulating first stars or clusters of galaxies.  The top-left panel
shows the ratio of $f_{l}/f_{l+1}$ which, remarkably, hovers around
the critical value of 0.0625 determined earlier.  This implies that
the relative cpu usage of each level, shown in the lower-left panel,
is nearly flat, and that going to additional levels of refinement only
increases the amount of calculation by a factor of roughly $1/l_{\rm
max}$.  Even in the most extreme case, the L80 run, the last level
adds only 25\% to the computation time.

The memory used by each level, on the other hand scales as:
\begin{equation}
{\rm memory} = C_2 \sum_l f_l N_{\rm root}^3 2^{3l}
\end{equation}
and the terms in this sum are shown in the bottom-right panel of
Figure~\ref{fig:scaling}.  This demonstrates that the memory usage is
dominated by the top three levels, and adding additional levels only
adds minimally to the memory usage of the run.  We have empirically
tested this scaling for cosmological simulations and found it to be
reasonably accurate (although the increase is typically slightly
larger than found here, usually 30-50\%, probably because of
less-than-ideal parallel scaling).

@@

These conclusions are, however, highly dependent on the type of run
being performed.  We contrast this cosmological case to the simulation
of a galaxy formation run, as shown by the dot-dashed curves in
Figure~\ref{fig:scaling}.  Note that the levels 1 and 2 are statically
refined to ensure the Lagrangian volume of the galaxy halo is resolved
to at least level 2 at all times. More importantly, the
$f_{l+1}/f_{l}$ ratio of the highest levels are systematically
slightly larger than the critical 0.0625 value, indicating that the
cpu time is dominated by the most refined levels, as shown in the
bottom-left panel.  The memory usage is dominated by levels 2 and 3,
the static levels, as the bottom-right panel demonstrates.  In this
case, adding more levels of refinement would somewhat increase the
cost of the simulation.

@@

Putting this together, we find an approximate scaling for the
computational time required for cosmological simulations:
\begin{equation}
t_{\rm SU} \propto M_{\rm res}^{-1} l_{\rm max}.
\end{equation}
This is approximately accurate, and gives users some way to estimate
compute times for cosmological calculations.  Similar scaling
relations can be obtained for other types of \enzo\ AMR simulations.

>>>>

The current version of \enzo\ has been parallelized for distributed
memory platforms using the Message Passing Interface (MPI).  This is
done using a single grid object as the basic unit of parallelization.
Each grid object -- including all cell and grid data -- is fully
contained in a single processor\footnote{In this context, we use the
word {\it processor} to mean a basic distribution unit; this could by
either a core or a node, depending on details of the system.  Note
that the current version of Enzo is not threaded, although the
development of a hybrid version is in process}.  Parallelization is
accomplished by distributing grids amongst processors.  This is done
on the root grid using a simple tiling system, where the root grid is
split up into $N_{\rm root}$ tiles, with $N_{\rm root}$ typically
equal to the number of processors $N_p$.

Load balancing on levels other than the root level (i.e., grids for
which the level $l > 0$) is different, as the refined patches are not
generally uniformly distributed.  Grids on refined levels are first
placed on the same processor as their parent to minimize
communication; however, this generally does not result in a
well-balanced computational load.  Therefore, the code has a number of options for
load balancing the grids on a given level. Each grid is assigned an
estimated computational load, generally equal to the number of cells in the grid (which,
empirically, is a good estimate of computational cost).  The first
load-balancing option is to move a grid from the processor with the
highest computational load to the processor with the lowest load, with the proviso
that only grids with load factors less than half the difference
between the highest and lowest loaded processors will be moved.  This
continues until the load ratio between the most-to-least loaded
processor is below 1.05 or until no suitable grid can be found to
transfer.  A second load balancing option uses a space-filling Hilbert
curve to order the grids by their approximate spatial position.  Then,
once the grids have a specific one-dimensional ordering, we can divide
up the grids into $N_p$ groups (with the division taking place as
equally as possible).  Load balancing is done separately for each
level.  Clearly load balancing is most successful if there are
significantly more grids than processors; however, small grids are
less efficient (because of their ghost zones), and so the code uses a
simple heuristic in order to split up grids until there are of order
10 grids per processor.  This generally results in good load-balancing
while not producing grids that are wastefully small.

@@

Because of the wide variety of simulations, methods, and uses of Enzo,
it is difficult to state in general terms which routines within the
code will be most costly during a given simulation.  As such, we have
designed a lightweight registering system that has been implemented
for the most commonly used routines (such as the hydrodynamic and
gravity solvers) as well as refinement level timers that measure the
time spent on each level.  Beyond this minimal set of routines, we
have designed a simple way for the user to modify the source by adding
\texttt{TIMER\_START(``Your_Routine_Name'')} and
\texttt{TIMER\_END(``Your_Routine_Name'')}.  These timers are created
and managed individually on each processor in an asynchronous fashion,
and contribute minimal computational, memory, and IO overhead.

At each complete root grid time step (or less often if specified),
each timer is then communicated to the root processor where it
calculates the mean, standard deviation, minimum, and maximum for each
of the timers across all processors.  For level timers, there are additional
attributes such as the number of cell updates, the current number of
grids, and the average cells/s/MPI process.  This information is then
output to a logfile.  This provides a simplified interface to the user
that can be used to diagnose performance issues as well as estimate a
given problem type's scalability.  In addition to the logfile, we have
developed a plotting interface for quickly producing figures that
process the data from the logfile.  These capabilities are described
in the online documentation, along with a further discussion of the
performance measurement implementation.

@@

In this section, we develop a simple, approximate model to estimate
the computational time and memory required to complete a given
calculation.  Note that we are {\it not} trying to model parallel
performance (which was discussed in the previous section), but have an
even simpler goal: to determine how to scale computational time
estimates as we increase the simulation resolution.  This is a
straightforward thing to do in codes that use static grids, as the
computational effort per time step is constant.  However, for an AMR
calculation, the number of grids that will be generated during the run
is not known in advance, and so the CPU time per problem time can vary
drastically throughout the calculation.  For example, at the beginning
of the simulation when the densities are nearly uniform, only the
static grid is required and the calculation progresses rapidly.
However, as structure forms and dense clumps are generated, the number
of grid points swells by orders of magnitude (an increase of $10^3$ is
not uncommon) and most of the CPU time is consumed at late times.
Therefore, simply performing a few steps at the beginning of the
calculation does produce a good estimate of the required CPU time.

@@

To make progress, we assume that the computational cost to advance a
single cell by one timestep is a constant.  This can be incorrect if
the chemistry solver requires many iterations, but is usually fairly
accurate.  For a unigrid calculation, the time would by proportional
to $N_{\rm root}^4$ since the number of cells scales as $N_{\rm
root}^3$ and, assuming the Courant condition is the 
factor that controls the time step, the number of steps to advance the
calculation over a given time interval is
proportional to $N_{\rm root}.$ Therefore, to advance a hierarchy a
given time interval, we find, accounting for all levels and using a refinement
factor of 2,
\begin{equation} t_{\rm SU} = C_1 \sum_l f_l N_{\rm root}^4 2^{4l}
\end{equation} where $C_1$ is a constant, which can be thought of as
the time taken to advance a single cell.  The factor $f_l$ is the
fraction of the volume on a given level that is actually refined.  By
definition $f_0 = 1$, and $f_l \le 1$.  In writing down this equation,
we neglect a number of costs that are not directly proportional to
cell count, including communication between processors, the $\log{N}$
factor for the root grid FFT, cache misses, optimizations, and other
costs associated with processing the hierarchy.  The first item on
this list, in particular, is clearly important for large processor
counts; however, we neglect parallel considerations in this section.

@@

\begin{figure}
\centerline{\includegraphics[width=0.7\textwidth]{figures/scaling_plot2}}
\caption{This figure show (clockwise from upper left), the covering
fraction $f_l$ of the grids on a given level $l$, the ratio of
$f_{l}/f_{l+1}$, the relative memory usage of each level (normalized
by the memory usage of level 0), and, in the lower-left panel, the
relative CPU usage of each level, again normalized by $l=0$.  In each
panel, three curves are shown: the solid black and long dashed red are
two cosmological simulations with box sizes of 30 h$^{-1}$ Mpc and
80h$^{-1}$ Mpc respectively.  The dot-dashed blue line is for a
non-cosmological simulation of a ram-pressure stripped galaxy.  In the
top-right panel, there is a line at the critical ratio of 0.06, which
determines if the next finer level ($l+1)$ takes more (above) or less
(below) CPU time than level $l$.}
\label{fig:scaling}
\end{figure}

As, the figure demonstrates, the $f_l$ values all show a sharp decline
with level, dropping as power-laws with the level $l$.  Focusing first
on the ``refine-everywhere" cosmological runs, which both show similar
behavior despite the different box sizes and redshifts at which we
collect the data.  In fact, we find these results are very typical for
cosmological runs, with only slight variations depending on the exact
system that is being simulated.  The top-left panel
shows the ratio of $f_{l}/f_{l+1}$ which, remarkably, hovers around
the critical value of 0.0625 determined earlier.  This implies that
the relative CPU usage of each level, shown in the lower-left panel,
is nearly flat, and that going to additional levels of refinement only
increases the amount of calculation by a factor of roughly $1/l_{\rm
max}$.  Even in the most extreme case, the L80 run, the last level
adds only 25\% to the computation time.

The memory used by each level, on the other hand, scales as
\begin{equation}
{\rm memory} = C_2 \sum_l f_l N_{\rm root}^3 2^{3l}
\end{equation}
and the terms in this sum are shown in the bottom-right panel of
Figure~\ref{fig:scaling}.  This demonstrates that the memory usage is
dominated by the top three levels, and adding additional levels only
adds minimally to the memory usage of the run.  We have empirically
tested this scaling for cosmological simulations and found it to be
reasonably accurate (although the increase is typically slightly
larger than found here, usually 30-50\%, probably because of
less-than-ideal parallel scaling).

@@

These conclusions are, however, highly dependent on the type of run
being performed.  We contrast this cosmological case to the simulation
of a galaxy formation run, as shown by the dot-dashed curves in
Figure~\ref{fig:scaling}.  Note that grid levels 1 and 2 in this simulation are statically
refined to ensure that the Lagrangian volume of the galaxy halo is resolved
to at least level 2 at all times. More importantly, the
$f_{l+1}/f_{l}$ ratio of the highest levels are systematically
slightly larger than the critical 0.0625 value, indicating that the
CPU time is dominated by the most refined levels, as shown in the
bottom-left panel.  The memory usage is dominated by levels 2 and 3,
the static levels, as the bottom-right panel demonstrates.  In this
case, adding more levels of refinement would somewhat increase the
cost of the simulation.

@@

Putting this together, we find an approximate scaling for the
computational time required for cosmological simulations:
\begin{equation}
t_{\rm SU} \propto M_{\rm res}^{-1} l_{\rm max}.
\end{equation}
This is approximately accurate, and gives users some way to estimate
compute times for \enzo\ calculations.
